---
layout: post
title: "2015-01-08-India-Education"
description: ""
category: R
tags: [rCharts, knitr]
---
{% include JB/setup %}

This is a quick example of producing a multiple bar chart using [rCharts]() package via implementing [HighCharts]() library. The original tutorial can be found in [rChart Gallery](http://rcharts.io/gallery/) but some functions are outdated as can be checked [here](https://github.com/ramnathv/rCharts/issues/251) - They are updated here.

Firstly _rCharts_ package needs to be installed from the GitHub repo if not installed.

```{r installPackages, eval=FALSE }
require(devtools)
install_github('rCharts', 'ramnathv')
```

The library is loaded.

```{r loadLibraries, message=FALSE, warning=FALSE}
library(rCharts)
library(knitr)
```

The data is shown below.

```{r prepareData, message=FALSE, warning=FALSE}
# Prepare data
x <- data.frame(USPersonalExpenditure)
colnames(x) <- substr(colnames(x), 2, 5)
kable(x)
```

The chart can be created as following. Note that the modified part is commented.

```{r createChart, message=FALSE, warning=FALSE}
# Create chart
a <- rCharts:::Highcharts$new() # modified
a$chart(type = "column")
a$title(text = "US Personal Expenditure")
a$xAxis(categories = rownames(x))
a$yAxis(title = list(text = "Billions of dollars"))
a$data(x)
```

Finally it is printed. Note that the `results` option should be set to be `'asis'`.

```{r printChart, message=FALSE, warning=FALSE, results='asis'}
# Print chart
a$print('chart 1', include_assets = TRUE) # modified
```