---
layout: post
title: "2016-05-13-Asynchronous-Processing-Using-Job-Queue"
description: ""
category: R
tags: [programming, jobqueue]
---

In this post, a way to overcome one of R's limitations (**lack of multi-threading**) is discussed by job queuing using the [jobqueue package](http://jobqueue.r-forge.r-project.org/) - a generic asynchronous job queue implementation for R. See the package description below.

> The jobqueue package is meant to provide an easy-to-use interface that allows to queue computations for background evaluation while the calling R session remains responsive. It is based on a *1-node socket cluster from the parallel package*. The package provides a way to do basic threading in R. The main focus of the package is on an intuitive and easy-to-use interface for the job queue programming construct. … Typical applications include: **background computation of lengthy tasks (such as data sourcing, model fitting, bootstrapping), simple/interactive parallelization (if you have 5 different jobs, move them to up to 5 different job queues), and concurrent task scheduling in more complicated R programs.** [Bramblecloud's cloud computing service for R](https://www.bramblecloud.com/) uses the jobqueue package as an asynchronous wrapper for synchronous remote connections.

Added to the typical applications indicated above, another good application of this package would be with a Shiny application especially when it is served by Shiny open source.

The package can be installed as following.

```{r load_libs, warning=FALSE, message=FALSE}
# http://r-forge.r-project.org/R/?group_id=2066
if(!require(jobqueue)) {
  pkg_src <- if(grepl("win", Sys.info()["sysname"], ignore.case = TRUE)) {
    "http://download.r-forge.r-project.org/bin/windows/contrib/3.2/jobqueue_1.0-4.zip"
  } else {
    "http://download.r-forge.r-project.org/src/contrib/jobqueue_1.0-4.tar.gz"
  }
  
  install.packages(pkg_src, repos = NULL)
}

library(jobqueue)
```

As can be seen in the description, it is highly related to the **parallel** package and thus it wouldn't be hard to understand how it works if you know how to make a cluster using that package - if not, have a look at [this post](http://jaehyeon-kim.github.io/2015/03/Parallel-Processing-on-Single-Machine-Part-I.html). 

```{r rf_ex, warning=FALSE}
# create queue
queue <- Q.make()
# load library
Q.push(queue, library(randomForest), mute = TRUE)
# execute function
Q.push(queue, randomForest(Species ~ ., data=iris, importance=TRUE, proximity=TRUE)$call)
# get output
Q.pop(queue)
# close queue
Q.close(queue)
```


```{r local_ex, warning=FALSE}
queue <- Q.make()
fun <- function(max_val) {
  unlist(lapply(1:max_val, function(x) {
    Sys.sleep(1)
    x
  }))
}
# send local R object
Q.sync(queue, fun)
Q.push(queue, fun(10))
Q.pop(queue)
while (TRUE) {
  out <- Q.pop(queue)
  message(paste("INFO output is NULL?", is.null(out)))
  if(!is.null(out)) {
    break
  }
}
Q.close(queue)
out
```






