---
layout: post
title: "2015-03-17-Parallel-Processing-on-Single-Machine-Part-II"
description: ""
category: R
tags: [foreach, doParallel, parallel, iterators, programming]
---
{% include JB/setup %}

In the [previous article](http://jaehyeon-kim.github.io/r/2015/03/14/Parallel-Processing-on-Single-Machine-Part-I/), parallel processing on a single machine using the **snow** and **parallel** packages are introduced. The four functions are an extension of `lapply()` with an additional argument that specifies a cluster object. In spite of their effectiveness and ease of use, there may be cases where creating a function that can be sent into clusters is not easy or looping may be more natural. In this article, another way of implementing parallel processing on a single machine is introduced using the **foreach** and **doParallel** packages where clusters using the socket transport are created using the **parallel** package. Finally the **iterators** package is briefly covered. The examples here are largely based on the individual packages' vignettes and further details can be found there.

Let's get started.

The following packages are loaded.

```{r load_pkgs, message=FALSE, warning=FALSE}
## load libraries
library(parallel)
library(iterators)
library(foreach)
library(doParallel)
```

A key difference between the *for* construct in base R and the *foreach* construct in the **foreach** package is

- *for* causes a *side-effect*
    + *side-effect* means state of something is changed. For example, printing a variable, changing the value of a variable and writing data to disk are side-effects.
- *foreach* returns a variable
    + By default, a list is created by a loop.

An equivalent outcome by the two can be created as following.

```{r eq_out, message=FALSE, warning=FALSE}
x = list()
for(i in 1:3) x[[length(x)+1]] = exp(i) # values of x is changed
x

x = foreach(i=1:3) %do% exp(i)
x
```

The *foreach* construct has two binary operators for executing a loop: `%do%` and `%dopar`. The first executes a loop sequentially while the latter does it in parallel. Note that, if a cluster object is not setup, the loop will be executed sequentially - the package prints a message for the first time only. 

```{r dos, message=FALSE, warning=FALSE}
system.time(foreach(i=1:4) %do% Sys.sleep(i))

cl = makeCluster(detectCores())
registerDoParallel(cl)
system.time(foreach(i=1:4) %dopar% Sys.sleep(i))
stopCluster(cl)
```

Multiple iterators can be used and (1) more than one expressions can be run in parentheses. (2) If more than one iterators are used, the number of iterations is the minimum length of the iterators. Some examples are shown below.

```{r basic, message=FALSE, warning=FALSE}
x = foreach(i=1:3, j=rep(10,3)) %do% (i + j)
x = foreach(i=1:3, j=rep(10,3)) %do% {
  # do something (1)
  i + j
}
x = foreach(i=rep(0,100), j=rep(10,3)) %do% (i + j)
do.call("sum",x) # (2)
```

There are some options to control a loop or to change the return data type. Some of them are

- `.combine`: A function can be specified to reduce the outcome variable. `c()`, `+ - * / ...`, `rbind/cbind` and `min()/max()` are some of useful built-in functions. A user-defined function can also be created.
    + `.multicombine` and `.maxcombine` can be set to determin how a function is applied - see the [vignette](http://cran.r-project.org/web/packages/foreach/vignettes/foreach.pdf) for further details.

```{r combine, message=FALSE, warning=FALSE}
foreach(i=1:2, .combine="c") %do% exp(i)
foreach(i=1:2, .combine="rbind") %do% exp(i)
foreach(i=1:2, .combine="+") %do% exp(i)
foreach(i=1:2, .combine="min") %do% exp(i)
# custom min function
cusMin = function(a, b) if(a < b) a else b
foreach(i=1:2, .combine="cusMin") %do% exp(i)
```

- `.inorder`: The order of sequence is reserved if *TRUE* and more relevant if a loop is executed in parallel. The default value is *TRUE*.

- `.packages`: By specifying one or more pckage names, the package(s) can be loaded in each cluster. This is similar to initialize a worker using `clusterEvalQ()` in the *parallel* package.

```{r pkgs, message=FALSE, warning=FALSE}
# return row dimension of Boston data in each cluster
cl = makeCluster(detectCores())
registerDoParallel(cl)
foreach(i=1:detectCores(), .combine="c", .packages="MASS") %dopar% dim(Boston)[1]
stopCluster(cl)
```

The binary operator of `%:%` can be used for list comprehension (filtering which to loop with `when`) and nested looping.

An example of list comprehension is shown below. It returns a vector of even numbers.

```{r lc, message=FALSE, warning=FALSE}
# reutrn a vector of even numbers
foreach(i=1:10, .combine="c") %:% when(i %% 2 == 0) %do% i
```

Below shows an example of nested looping by *for* and *foreach*. According to the [vignette](http://cran.r-project.org/web/packages/foreach/vignettes/foreach.pdf), it is not necessary to determine which loop (inner or outer) to parallize as `%:%` turns multiple foreach loops into a single stream of tasks that can be parallelized.

```{r nesting, message=FALSE, warning=FALSE}
# for
avec = c(10,20)
bvec = 1:4
mat = matrix(0, nrow=length(avec), ncol=length(bvec))
for(b in bvec) {
  for(a in 1:length(avec)) {
    mat[a,b] = avec[a] + bvec[b]
  }
}
mat

# foreach - sequential
foreach(b=1:4, .combine="cbind") %:%
  foreach(a=c(10,20), .combine="c") %do% (a + b)

# foreach - parallel
cl = makeCluster(detectCores())
registerDoParallel(cl)
foreach(b=1:4, .combine="cbind") %:%
  foreach(a=c(10,20), .combine="c") %dopar% (a + b)
stopCluster(cl)
```

As a loop is constructed by *foreach*, the **iterators** package can be useful to be used together as the package allows to create an iterator object from a conventional R objects: vector, data.frame and even function. Some examples are shown below.

```{r iter, message=FALSE, warning=FALSE}
## create iterator object
# by object
iters = iter(1:10)
c(nextElem(iters),nextElem(iters))

df = data.frame(number=c(1:26), letter=letters)
iters = iter(df, by="row")
nextElem(iters)

iters = iter(list(1:2, 3:4))
for(i in 1:iters$length) print(nextElem(iters))

# by function
set.seed(1237)
iters = iter(function() sample(0:9, 4, replace=TRUE))
nextElem(iters)
```

The following code generates *Error: StopIteration* error at the end as there is no *nextElem* available.

```{r err, message=FALSE, warning=FALSE, eval=FALSE}
iters = iter(list(1:2, 3:4))
for(i in 1:(iters$length+1)) print(nextElem(iters))
```

A quick example of using an iterator object with *foreach* is shown below.

```{r ex, message=FALSE, warning=FALSE, eval=FALSE}
set.seed(1237)
mat = matrix(rnorm(100),nrow=1)
iters = iter(mat, by="row")
foreach(a=iters, .combine="c") %do% mean(a)
```

Finally this package provides some wappers around some of built-in functions: `icount()`, `irnorm()`, `irunit()`, `irbinom()`, `irnbinom()` and `irpois()`.




