---
layout: post
title: "2015-01-14-Quick-Trial-to-Add-Column"
description: ""
category: R
tags: [plyr, dplyr, data.table]
---
{% include JB/setup %}

This is a quick trial of adding overall and conditional (by user) average columns in a data frame. `base`,`plyr`,`dplyr`,`data.table`,`dplyr + data.table` packages are used. Personally I perfer `dplyr + data.table` - `dplyr` for comperhensive syntax and `data.table` for speed.

```{r setup, warning=FALSE, message=FALSE}
## set up variables
size <- 36000
numUsers <- 4900
# roughly each user has 7 sessions
numSessions <- (numUsers / 7) - ((numUsers / 7) %% 1)
 
## create data frame
set.seed(123457)
userIds <- sample.int(numUsers, size=size, replace=TRUE)
ssIds <- sample.int(numSessions, size=size, replace=TRUE)
scores <- sample.int(10, size=size, replace=TRUE) 

preDf <- data.frame(User=userIds, Session=ssIds, Score=scores)
preDf$User <- as.factor(preDf$User) 
```

#### Adding overall average

As calculating overall average is not complicated, I don't find a big difference among packages.

```{r base_overall, message=FALSE, warning=FALSE}
# base
system.time(overallDf1 <- transform(preDf, MeanScore=mean(Score, na.rm=TRUE))) 
```

```{r plyr_overall, message=FALSE, warning=FALSE}
# plyr
require(plyr)
system.time(overallDf2 <- mutate(preDf, MeanScore=mean(Score, na.rm=TRUE)))
```

```{r dplyr_overall, message=FALSE, warning=FALSE}
# dplyr
require(dplyr)
system.time(overallDf3 <- preDf %>% 
                            mutate(MeanScore=mean(Score, na.rm=TRUE)))
```

```{r dt_overall, message=FALSE, warning=FALSE}
# data.table
require(data.table)
preDt <- data.table(preDf)
setkey(preDt, User)
system.time(overallDt <- preDt[,list(User=User
                                     ,Session=Session
                                     ,Score=Score
                                     ,MeanScore=mean(Score, na.rm=T))])
```

```{r combined_overall, message=FALSE, warning=FALSE}
# dplyr + data.table
system.time(overallDf4 <- preDt %>% 
                            mutate(MeanScore=mean(Score, na.rm=TRUE))) 
```

#### Adding average by user

It takes quite long using `plyr` and other packages would be more practial - `base` is not considered even.

```{r plyr_con, message=FALSE, warning=FALSE}
# plyr
require(plyr)
system.time(postDf1 <- ddply(preDf
                             ,.(User)
                             ,mutate,MeanScore=mean(Score, na.rm=TRUE))) 
```

```{r dplyr_con, message=FALSE, warning=FALSE}
# dplyr
require(dplyr)
system.time(postDf2 <- preDf %>% 
              group_by(User) %>% 
              mutate(MeanScore=mean(Score, na.rm=TRUE)) %>% 
              arrange(User))
```

```{r dt_con, message=FALSE, warning=FALSE}
# data.table
require(data.table)
preDt <- data.table(preDf)
setkey(preDt, User)
system.time(postDt <- preDt[,list(Session=Session
                                  ,Score=Score
                                  ,MeanScore=mean(Score, na.rm=T))
                            ,by=User])
```

```{r combined_con, message=FALSE, warning=FALSE}
# dplyr + data.table
system.time(postDf3 <- preDt %>% 
              group_by(User) %>% 
              mutate(MeanScore=mean(Score, na.rm=TRUE)) %>% 
              arrange(User))
```
